Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers Across Aurignacian Time Periods

### 2025-02-18 endofquarterpaper.qmd
permanova pipeline
extra

---
title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
zotero: "Signbase-study"
link-citations: true
---

@@

A distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019]. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe [@Boyd_Silk_2020]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian period. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016]. In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

@@

Similar analyses (seriation, correspondence and GIS) were also performed upon personal ornament types from the Aurignacian period as a proxy for linguistic diversity. The data set consisted of 157 distinct ornament types, and the analyses identified fourteen geographically cohesive clusters sharing similar ornament type associations [@Vanhaeren_d]. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. Availability of raw materials was considered ~in regards to~ the potential explanation that observed patterns are determined by ~such~. ~It was concluded that, due to observed patterns in the data that countered that explanation~, [BM: They found that] the differences in personal ornamentation could not be explained by material availability.

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

@@

In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects. Firstly, we identify clusters using seriation analysis, network analysis, and PCoA. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.

@@

Principal coordinate decomposition analysis (PCoA) analyzes the eigenvalue decomposition of the Jaccard matrix and arranges the sites by dissimilarity [@Legendre_Legendre_2012]. This provides a visual of sign similarity as visualized by spacing and arrangement. The PCoA in this paper is calibrated to be plotted on a scatterplot, as opposed to a biplot.

>>>>

---
title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers Across Aurignacian Time Periods"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
zotero: "Signbase-study"
link-citations: true
---

@@

A distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019]. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe [@Boyd_Silk_2020]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].The Aurignacian can be divided up into four chronological periods: Transitional, Proto-Aurignacian, Early Aurignacian, and Evolved Aurignacian [@Banks_dErrico_Zilhao_2013, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021].

In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the subsections of the Aurignacian period. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016]. In this study, we address the following questions: How does sign distribution change across periods of the Aurignacian, what do the patterns tell us about cultural diversity patterns and movement of Upper Paleolithic humans throughout the Aurignacian?

@@

Similar analyses (seriation, correspondence and GIS) were also performed upon personal ornament types from the Aurignacian period as a proxy for linguistic diversity. The data set consisted of 157 distinct ornament types, and the analyses identified fourteen geographically cohesive clusters sharing similar ornament type associations [@Vanhaeren_d]. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. The possibility that observed patterns were influenced by availability of raw materials was considered, and ultimately discarded as an explanation for the patterns.

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility and comparison of in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

@@

In this paper, we test the hypothesis that cultural groups are discernible in a data set of geometric signs on mobile objects. Firstly, we identify clusters within each time period using seriation and network analysis. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance,

@@
